disaggregation: An \proglang{R} package for Bayesian spatial disaggregation modelling

### 2019-11-18 article.Rnw
Add caption to table
Minor update to abstract

%% - \Abstract{} almost as usual
\Abstract{
  Disaggregation modelling, or downscaling, has been used across many fields including epidemiology and ecology. These models use repsonse data agreggated of large heterogenous regions and predict the response at fine-scale over the region by using fine-scale covariates to inform the heterogeneity. This paper presents the \proglang{R} package \emph{disaggregation}, which provides the functionality to streamline the process of running a disaggregation model for fine scale predictions.
}

@@

Methods for estimating high resolution repsonse information from models of aggregated repsonse data over large spatial regions are 
becoming increasingly sought after in disease risk~\cite{li2012log, diggle2013spatial}, expecially in malaria mapping~\cite{sturrock2014fine, alegana2016advances}, and ecological speices distribution~\cite{keil2013downscaling, barwell2014can}. The aggreagtion of repsonse data over large 
heterongenous regions is problematic for understanding the fine scale distribution. However, by using fine scale information from related covariates we can inform the hetrogeneity of the response variable of interest within the agreggated area. This method has recently been applied in the latest global high-resolution maps of \emph{Plasmodium falciparum} and \emph{Plasmodium vivax} malaria~\cite{falciparum2019, vivax2019}.

@@

Let's say we have response data for $N$ polygons, which corresponds to count data for the property of interest within that polygon, $y_j$. By count data here we mean the data at the level that can be aggregated by simple addition, i.e. the count data of the polygon is given by the sum of the count data for all the pixels within that polygon. The rate, $p_i$ is defined such that the number of cases in a pixel can be calculated by multiplying the rate by the aggregation raster. For example, in epidemiology, we may have as our response the number of people with a certain disease (case incidence). Our rate would be the incidence rate as a fraction of the population, where the aggregation raster corresponds to population.

@@

The model described in Section~\ref{sec:code} for malaria in Madagascar has been optimised using the \pkg{disaggregation} package. Here we fit the same model by running MCMC using the \pkg{tmbstan} package, with the NUTS algorithm using four chains. It is important to note that this is a useful feature of the \pkg{disaggregation} package, to be able to create the \pkg{TMB} model object (using the \code{fit_model} function with one iteration) and pass it directly to \pkg{tmbstan}. The model is fitted by running the MCMC algorithm for 8000 iterations with 2000 of those as warmup, which took 94 hours. This number of iterations was chosen by running the MCMC algorithm repeatedly, starting at 1000 iterations and doubling the number of iterations each time, until the value of the MCMC convergence statistic, $\hat{R}$, dropped below 1.05 for all model parameters. The model is optimised using \pkg{TMB} within the \pkg{disaggregation} package took 56 seconds. Fitted parameter values for both of these methods are given in the table below.

@@

@@

%
\begin{table}
\centering
\begin{tabular}{l|ll|ll}
 & \multicolumn{2}{|c|}{MCMC (94 hours)} & \multicolumn{2}{|c}{TMB (56 seconds)} \\
Parameter & Mean & SD & Mean & SD \\
\hline
Intercept & -3.09 & 0.33 & -3.12 & 0.27 \\
Slope 1 & 0.20 & 0.28 & 0.24 & 0.27 \\
Slope 2 & 0.32 & 0.21 & 0.30 & 0.21\\
Slope 3 & -0.36 & 0.20 & -0.37 & 0.20 \\
Slope 4 & -0.24 & 0.23 & -0.24 & 0.21 \\
iid $log(\tau)$ & 1.08 & 0.27 & 1.10 & 0.27 \\
$log(\sigma)$ & -3.34 & 0.59 & -3.55 & 0.59 \\
$log(\rho)$ & 0.72 & 0.31 & -0.56 & 0.30 \\
\end{tabular}
\end{table}
%

>>>>

%% - \Abstract{} almost as usual
\Abstract{
  Disaggregation modelling, or downscaling, has been used across many fields including epidemiology and ecology. In these fields surveillance data, aggregated over large regions, is becoming more common, leading to an increasing demand for modelling frameworks that can deal with this data to understand spatial patterns. Disaggregation regression models use repsonse data agreggated over large heterogenous regions and predict the response at fine-scale over the region by using fine-scale covariates to inform the heterogeneity. This paper presents the \proglang{R} package \emph{disaggregation}, which provides the functionality to streamline the process of running a disaggregation model for fine scale predictions.
}

@@

Methods for estimating high resolution repsonse information from models of aggregated repsonse data over large spatial regions are 
becoming increasingly sought after in disease risk~\cite{li2012log, diggle2013spatial}, especially in malaria mapping~\cite{sturrock2014fine, alegana2016advances}, and ecological speices distribution~\cite{keil2013downscaling, barwell2014can}. The aggreagtion of repsonse data over large 
heterongenous regions is problematic for understanding the fine scale distribution. However, by using fine scale information from related covariates we can inform the hetrogeneity of the response variable of interest within the agreggated area. This method has recently been applied in the latest global high-resolution maps of \emph{Plasmodium falciparum} and \emph{Plasmodium vivax} malaria~\cite{falciparum2019, vivax2019}.

@@

Suppose we have response data for $N$ polygons, which corresponds to count data for the property of interest within that polygon, $y_j$. By count data here we mean the data at the level that can be aggregated by simple addition, i.e. the count data of the polygon is given by the sum of the count data for all the pixels within that polygon. The rate, $p_i$, is defined such that the number of cases in a pixel can be calculated by multiplying the rate by the aggregation raster. For example, in epidemiology, we may have as our response the number of people with a certain disease (case incidence). Our rate would be the incidence rate as a fraction of the population, where the aggregation raster corresponds to population.

@@

The model described in Section~\ref{sec:code} for malaria in Madagascar has been optimised using the \pkg{disaggregation} package. Here we fit the same model by running MCMC using the \pkg{tmbstan} package, with the NUTS algorithm using four chains. It is important to note that this is a useful feature of the \pkg{disaggregation} package, to be able to create the \pkg{TMB} model object (using the \code{fit_model} function with one iteration) and pass it directly to \pkg{tmbstan}. The model is fitted by running the MCMC algorithm for 8000 iterations with 2000 of those as warmup, which took 94 hours. This number of iterations was chosen by running the MCMC algorithm repeatedly, starting at 1000 iterations and doubling the number of iterations each time, until the value of the MCMC convergence statistic, $\hat{R}$, dropped below 1.05 for all model parameters. The model is optimised using \pkg{TMB} within the \pkg{disaggregation} package took 56 seconds. Fitted parameter values for both of these methods are given in Table~\ref{tab:mcmcresults}.

@@

%
\begin{table}
\centering
\begin{tabular}{l|ll|ll}
 & \multicolumn{2}{|c|}{MCMC (94 hours)} & \multicolumn{2}{|c}{TMB (56 seconds)} \\
Parameter & Mean & SD & Mean & SD \\
\hline
Intercept & -3.09 & 0.33 & -3.12 & 0.27 \\
Slope 1 & 0.20 & 0.28 & 0.24 & 0.27 \\
Slope 2 & 0.32 & 0.21 & 0.30 & 0.21\\
Slope 3 & -0.36 & 0.20 & -0.37 & 0.20 \\
Slope 4 & -0.24 & 0.23 & -0.24 & 0.21 \\
iid $log(\tau)$ & 1.08 & 0.27 & 1.10 & 0.27 \\
$log(\sigma)$ & -3.34 & 0.59 & -3.55 & 0.59 \\
$log(\rho)$ & 0.72 & 0.31 & -0.56 & 0.30 \\
\end{tabular}
\caption{Fitted model parameter values using both MCMC and using \pkg{TMB} within the \pkg{disaggregation} package.}
\label{tab:mcmcresults}
\end{table}
%

@@
